decompr: GVC decomposition in R

### 2015-01-22 draft.Rnw
Update draft.Rnw
Shift alignment tab in Leontief equation
Update draft.Rnw
Various changes pt.I

\vspace{\stretch{1}}
\abstract{The R package \textit{decompr} implements algorithms for the analysis of Global Value Chains.
Two methods for gross export flow decomposition using Inter-Country Input-Output tables are provided.
The first method is the \textit{Wang-Wei-Zhu} algorithm which splits bilateral gross exports into 16 value added components.
These components can broadly be divided into domestic and foreign value added.
The second method concerns a decomposition based on the classical \textit{Leontief} insight.
This derives the value added origin of a country-industrys exports by source country and source industry,
by applying the basic Leontief insight to gross trade data.
This article summarises the methodology of the algorithms and describes the format of the input and output data.}
\pagebreak

\section{Introduction}
Global Value Chains (GVCs) have become a central factor in trade and development policy.
Policy makers from different countries and institutions have placed them at the center of their agenda
and continuously emphasize their growing importance for both international trade and economic development.
However, analysing this phenomenon empirically requires complex matrix manipulations,
since the relevant data is only available in the form of gross flows.
This package enables researchers with little background in matrix algebra and linear programming
to easily derive standard GVC indicators for statistical analysis.

As mentioned above, \textit{decompr} uses Inter-Country Input-Output tables,
such as those published by the OECD, WIOD, or national statistics bureaus.
These tables state input-output relationships in gross terms
between industries within and across countries.

@@

The source decomposition of gross trade flows solves this problem,
by reallocating the value of intermediate goods used by industries to the original producers.
In our example, the use of Argentinian agricultural produce (raw hides)
is subtracted from the Turkish leather industry and added to the Argintinian agricultural industry.

The Wang-Wei-Zhu (henceforth WWZ) decomposition goes a step further by not only revealing the source of the value added,
but also breaking down exports into different categories, based on its usage and final destination.

@@

\section{Data}
Two data sets are included in the package.
The WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world}. 
Furthermore, a fictional 3-country 3-sector data set is included, which we will use throughout this article to exemplify the mechanics of the package.

@@

@@

@@

\begin{align}
\begin{split}
V (I-A)^{-1}E=&
\begin{pmatrix}
v_{i}^{k}& 0& 0& 0\\
0& v_{j}^{k}& 0& 0\\
0& 0& v_{i}^{l}& 0\\
0& 0& 0& v_{j}^{l}
\end{pmatrix}
*
\begin{pmatrix}
b_{ii}^{kk}& b_{ij}^{kk}& b_{ii}^{kl}& b_{ij}^{kl}\\
b_{ji}^{kk}& b_{jj}^{kk}& b_{ji}^{kl}& b_{jj}^{kl}\\
b_{ii}^{lk}& b_{ij}^{lk}& b_{ii}^{ll}& b_{ij}^{ll}\\
b_{ji}^{lk}& b_{jj}^{lk}& b_{ji}^{ll}& b_{jj}^{ll}
\end{pmatrix}
*
\begin{pmatrix}
e_{i}^{k}& 0& 0& 0\\
0& e_{j}^{k}& 0& 0\\
0& 0& e_{i}^{l}& 0\\
0& 0& 0& e_{j}^{l}
\end{pmatrix}
 \\
=& 
\begin{pmatrix}
v_{i}^{k}b_{ii}^{kk}e_{i}^{k}& v_{i}^{k} b_{ij}^{kk}e_{j}^{k}& v_{i}^{k}b_{ii}^{kl}e_{i}^{l}& v_{i}^{k}b_{ij}^{kl}e_{j}^{l}\\
v_{j}^{k}b_{ji}^{kk}e_{i}^{k}& v_{j}^{k}b_{jj}^{kk}e_{j}^{k}& v_{j}^{k}b_{ji}^{kl}e_{i}^{l}& v_{j}^{k}b_{jj}^{kl}e_{j}^{l}\\
v_{i}^{l}b_{ii}^{lk}e_{i}^{k}& v_{i}^{l}b_{ij}^{lk}e_{j}^{k}& v_{i}^{l}b_{ii}^{ll}e_{i}^{l}& v_{i}^{l}b_{ij}^{ll}e_{j}^{l}\\
v_{j}^{l}b_{ji}^{lk}e_{i}^{k}& v_{j}^{l}b_{jj}^{lk}e_{j}^{k}& v_{j}^{l}b_{ji}^{ll}e_{i}^{l}& v_{j}^{l}b_{jj}^{ll}e_{j}^{l}
\end{pmatrix}
=
\begin{pmatrix}
vax_{ii}^{kk}& vax_{ij}^{kk}& vax_{ii}^{kl}& vax_{ij}^{kl}\\
vax_{ji}^{kk}& vax_{jj}^{kk}& vax_{ji}^{kl}& vax_{jj}^{kl}\\
vax_{ii}^{lk}& vax_{ij}^{lk}& vax_{ii}^{ll}& vax_{ij}^{ll}\\
vax_{ji}^{lk}& vax_{jj}^{lk}& vax_{ji}^{ll}& vax_{jj}^{ll}
\end{pmatrix}
\end{split}
\end{align}
with
\begin{align*}
&v_{s}^{c} = \frac{va_{s}^{c}}{y_{s}^{c}} = 1 - \sum_{d}^{2}{a_{kc}^{ds}} - \sum_{d}^{2}{a_{lc}^{ds}} \hspace{2cm} (c \in k,l \hspace{.5cm} s \in i,k),\\
&\begin{pmatrix}
b_{ii}^{kk}& b_{ij}^{kk}& b_{ii}^{kl}& b_{ij}^{kl}\\
b_{ji}^{kk}& b_{jj}^{kk}& b_{ji}^{kl}& b_{jj}^{kl}\\
b_{ii}^{lk}& b_{ij}^{lk}& b_{ii}^{ll}& b_{ij}^{ll}\\
b_{ji}^{lk}& b_{jj}^{lk}& b_{ji}^{ll}& b_{jj}^{ll}
\end{pmatrix}
=
\begin{pmatrix}
1-a_{ii}^{kk}& -a_{ij}^{kk}& -a_{ii}^{kl}& -a_{ij}^{kl}\\
-a_{ji}^{kk}& 1-a_{jj}^{kk}& -a_{ji}^{kl}& -a_{jj}^{kl}\\
-a_{ii}^{lk}& -a_{ij}^{lk}& 1-a_{ii}^{ll}& -a_{ij}^{ll}\\
-a_{ji}^{lk}& -a_{jj}^{lk}& -a_{ji}^{ll}& 1-a_{jj}^{ll}
\end{pmatrix}^{-1},
\end{align*}
and
\begin{equation*}
a_{cf}^{su} = \frac{inp_{cf}^{su}}{y_{f}^{u}}  \hspace{2cm} (c,f \in k,l \hspace{.5cm} s,u \in i,k),
\end{equation*}
where $v_{s}^{c}$ gives the share of value added, $va_{s}^{c}$, in output , $y_{s}^{c}$, and $e_{s}^{c}$ indicates gross exports. Finally, $a_{cf}^{su}$ denotes the share of inputs , $inp_{cf}^{su}$, in output. The elements of the $V(I-A)^{-1}E$ or $vax$ matrix are our estimates for the country-industry level value added origins of each country-industry's exports. decompr implements this algorithm into R to automate the process of deriving the matrix. Equipped with it, researchers can calculate standard GVC indicators. Examples include \citet{dahuetal01}'s Vertical Specialisation ratio at the industry-level which can be derived by summing for each country and industry across the value added of all foreign countries and industries. Alternatively, the four dimensions of the matrix (source country, source industry, using country, using industry) allow for industry-level gravity-type estimations of value added trade flows.

>>>>

\vspace{\stretch{1}}
\abstract{Global Value Chains have become a central unit of analysis in research on international trade. However, 
the complex matrix transformations at the basis of most Value Chain indicators still constitute a significant entry barrier
the field. The R package \textit{decompr} solves this problem by implementing algorithms for the analysis of Global Value Chains,
which calculate standard indicators in an automated fashion. Two methods for gross export flow decomposition using 
Inter-Country Input-Output tables are provided. The first method is the \textit{Wang-Wei-Zhu} algorithm which splits 
bilateral gross exports into 16 value added components. These components can broadly be divided into domestic and 
foreign value added. The second method concerns a decomposition based on the classical \citet{wale36} insight.
It derives the value added origins of an industry's exports by source country and source industry, using easily available
gross trade data. This article summarises the methodology of the algorithms, describes the format of the input and output data, 
and exemplifies the usefulness of the two methods on the basis of a simple example data set.}
\pagebreak

\section{Introduction}
The analysis of Global Value Chains (GVCs) has become a central topic in research on trade and development policy.
Both policy makers and academia increasingly value the growth opportunities GVCs offer to world trade and, especially, 
to developing countries. However, analysing this phenomenon empirically requires complex matrix manipulations,
since the relevant data is only available in the form of gross flows.
This package enables researchers with little background in matrix algebra and linear programming
to easily derive standard GVC indicators for statistical analysis.

\textit{decompr} uses Inter-Country Input-Output tables (ICIOs), such as those published by the OECD, 
the World Input Output Database (WIOD),  or national statistics bureaus, as input. 
These tables state supply and demand relationships in gross terms between industries within and across countries.

@@

The \textit{Leontief} decomposition of gross trade flows solves this problem
by reallocating the value of intermediate goods used by industries to the original producers.
In our example, the use of Argentinian agricultural produce (raw hides)
is subtracted from the Turkish leather industry and added to the Argentinian agricultural industry.

The \textit{Wang-Wei-Zhu} (henceforth WWZ) decomposition goes a step further by not only revealing the source of the value added,
but also breaking down exports into different categories, based on its final usage and destination. It implements the 
theoretical work of \citet{wang2014quantifying}. The main categories are:

@@

\section{Data}
Two data sets are included in the package, namely
the WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world} and, 
furthermore, a fictional 3-country 3-sector data set is included, which we will use throughout this article to exemplify
the mechanics of the package.

@@

The data is set up in order to illustrate the benefits of the decompositions. That is, we follow the flows of intermediate
goods through a fictional GVC and show how the readily available gross trade flows differ from the decomposed value added
flows. To this end, we construct the elements of the input-output tables such that we have two countries and two industries
that focus on upstream tasks, which means they focus on supplying other industries, and one industry and one country that
is specialized in downstream tasks, that is it serves mainly final demand. In our example the upstream industries are 
Agriculture and Textiles while the downstream industry in Transport Equipment. Similarly, Argentina and Turkey represent 
upstream countries with Germany being located downstream within this specific value chain.

@@

Let us now turn to the actual algorithms starting with the \textit{Leontief} decomposition. We shortly describe the theoretical
derivation of the method to expose the inner steps of the \textit{decompr} package. Then we turn to the technical implementation
and finish with a description of the output.

@@

\begin{align}
\begin{split}
V (I- & A)^{-1} E =
\begin{pmatrix}
v_{i}^{k}& 0& 0& 0\\
0& v_{j}^{k}& 0& 0\\
0& 0& v_{i}^{l}& 0\\
0& 0& 0& v_{j}^{l}
\end{pmatrix}
*
\begin{pmatrix}
b_{ii}^{kk}& b_{ij}^{kk}& b_{ii}^{kl}& b_{ij}^{kl}\\
b_{ji}^{kk}& b_{jj}^{kk}& b_{ji}^{kl}& b_{jj}^{kl}\\
b_{ii}^{lk}& b_{ij}^{lk}& b_{ii}^{ll}& b_{ij}^{ll}\\
b_{ji}^{lk}& b_{jj}^{lk}& b_{ji}^{ll}& b_{jj}^{ll}
\end{pmatrix}
*
\begin{pmatrix}
e_{i}^{k}& 0& 0& 0\\
0& e_{j}^{k}& 0& 0\\
0& 0& e_{i}^{l}& 0\\
0& 0& 0& e_{j}^{l}
\end{pmatrix}
=
\\
&
\begin{pmatrix}
v_{i}^{k}b_{ii}^{kk}e_{i}^{k}& v_{i}^{k} b_{ij}^{kk}e_{j}^{k}& v_{i}^{k}b_{ii}^{kl}e_{i}^{l}& v_{i}^{k}b_{ij}^{kl}e_{j}^{l}\\
v_{j}^{k}b_{ji}^{kk}e_{i}^{k}& v_{j}^{k}b_{jj}^{kk}e_{j}^{k}& v_{j}^{k}b_{ji}^{kl}e_{i}^{l}& v_{j}^{k}b_{jj}^{kl}e_{j}^{l}\\
v_{i}^{l}b_{ii}^{lk}e_{i}^{k}& v_{i}^{l}b_{ij}^{lk}e_{j}^{k}& v_{i}^{l}b_{ii}^{ll}e_{i}^{l}& v_{i}^{l}b_{ij}^{ll}e_{j}^{l}\\
v_{j}^{l}b_{ji}^{lk}e_{i}^{k}& v_{j}^{l}b_{jj}^{lk}e_{j}^{k}& v_{j}^{l}b_{ji}^{ll}e_{i}^{l}& v_{j}^{l}b_{jj}^{ll}e_{j}^{l}
\end{pmatrix}
=
\begin{pmatrix}
vax_{ii}^{kk}& vax_{ij}^{kk}& vax_{ii}^{kl}& vax_{ij}^{kl}\\
vax_{ji}^{kk}& vax_{jj}^{kk}& vax_{ji}^{kl}& vax_{jj}^{kl}\\
vax_{ii}^{lk}& vax_{ij}^{lk}& vax_{ii}^{ll}& vax_{ij}^{ll}\\
vax_{ji}^{lk}& vax_{jj}^{lk}& vax_{ji}^{ll}& vax_{jj}^{ll}
\end{pmatrix}
\end{split}
\end{align}
with
\begin{align*}
&v_{s}^{c} = \frac{va_{s}^{c}}{y_{s}^{c}} = 1 - \sum_{d}^{2}{a_{kc}^{ds}} - \sum_{d}^{2}{a_{lc}^{ds}} \hspace{2cm} (c \in k,l \hspace{.5cm} s \in i,k),\\
&\begin{pmatrix}
b_{ii}^{kk}& b_{ij}^{kk}& b_{ii}^{kl}& b_{ij}^{kl}\\
b_{ji}^{kk}& b_{jj}^{kk}& b_{ji}^{kl}& b_{jj}^{kl}\\
b_{ii}^{lk}& b_{ij}^{lk}& b_{ii}^{ll}& b_{ij}^{ll}\\
b_{ji}^{lk}& b_{jj}^{lk}& b_{ji}^{ll}& b_{jj}^{ll}
\end{pmatrix}
=
\begin{pmatrix}
1-a_{ii}^{kk}& -a_{ij}^{kk}& -a_{ii}^{kl}& -a_{ij}^{kl}\\
-a_{ji}^{kk}& 1-a_{jj}^{kk}& -a_{ji}^{kl}& -a_{jj}^{kl}\\
-a_{ii}^{lk}& -a_{ij}^{lk}& 1-a_{ii}^{ll}& -a_{ij}^{ll}\\
-a_{ji}^{lk}& -a_{jj}^{lk}& -a_{ji}^{ll}& 1-a_{jj}^{ll}
\end{pmatrix}^{-1},
\end{align*}
and
\begin{equation*}
a_{cf}^{su} = \frac{inp_{cf}^{su}}{y_{f}^{u}}  \hspace{2cm} (c,f \in k,l \hspace{.5cm} s,u \in i,k),
\end{equation*}
where $v_{s}^{c}$ gives the share of value added, $va_{s}^{c}$, in output , $y_{s}^{c}$, and $e_{s}^{c}$ indicates gross exports. Finally, $a_{cf}^{su}$ denotes the share of inputs , $inp_{cf}^{su}$, in output. The elements of the $V(I-A)^{-1}E$ or $vax$ matrix are our estimates for the country-industry level value added origins of each country-industry's exports. decompr implements this algorithm into R to automate the process of deriving the matrix. Equipped with it, researchers can calculate standard GVC indicators. Examples include \citet{dahuetal01}'s Vertical Specialisation ratio at the industry-level which can be derived by summing for each country and industry across the value added of all foreign countries and industries. Alternatively, the four dimensions of the matrix (source country, source industry, using country, using industry) allow for industry-level gravity-type estimations of value added trade flows.
